---
title: "Class 09: Candy Mini Project"
author: "Julianne Lee (PID: A17979401)"
format: pdf
toc: true
---

## Background

Today we are taking a detor to analyze a dataset (that we have more intrinsic insight into) with the most useful analysis method we have learned thus far - Principal Component Analysis (PCA)

## Data import
The data is all related to halloween candy and is from the 538 website.

```{r}
candy <- read.csv("candy-data.txt", row.names =1)
head(candy)
```

> Q1. How many different candy types are in this dataset?

```{r}
nrow(candy)
```

> Q2. How many fruity candy types are in the dataset?

```{r}
sum(candy$fruity)
```

> Q3. What is your favorite candy (other than Twix) in the dataset and what is it’s winpercent value?

```{r}
library(dplyr)
candy |> 
  filter(row.names(candy)=="Nerds") |> 
  select(winpercent)
```

> Q4. What is the winpercent value for “Kit Kat”?

```{r}
candy |> 
  filter(row.names(candy)=="Kit Kat") |> 
  select(winpercent)
```

> Q5. What is the winpercent value for “Tootsie Roll Snack Bars”?

```{r}
candy |> 
  filter(row.names(candy)=="Tootsie Roll Snack Bars") |> 
  select(winpercent)
```

> Q6. Is there any variable/column that looks to be on a different scale to the majority of the other columns in the dataset?
>
> > winpercent

```{r}
library("skimr")
skim(candy)
```

> Q7. What do you think a zero and one represent for the candy$chocolate column?
>
> > 0: not chocolate, 1: yes chocolate

## Exploratory analysis

> Q8. Plot a histogram of winpercent values using both base R and ggplot2.

```{r}
library(ggplot2)

hist(candy$winpercent)

ggplot(candy, aes(winpercent)) + geom_histogram(bins=20)
```

> Q9. Is the distribution of winpercent values symmetrical?
>
> > No

> Q10. Is the center of the distribution above or below 50%?

```{r}
median(candy$winpercent)
summary(candy$winpercent)
```

> Q11. On average is chocolate candy higher or lower ranked than fruit candy?

```{r}
choc <- candy[as.logical(candy$chocolate),]
fruit <- candy[as.logical(candy$fruity),]
mean(choc$winpercent) > mean(fruit$winpercent)
```

> Q12. Is this difference statistically significant?

```{r}
t.test(choc$winpercent,fruit$winpercent)
```

## Overall Candy Rankings

> Q13. What are the five least liked candy types in this set?

```{r}
candy |> arrange(winpercent) |> head(5)
```

> Q14. What are the top 5 all time favorite candy types out of this set?

```{r}
candy |> arrange(desc(winpercent)) |> head(5)
```

> Q15. Make a first barplot of candy ranking based on winpercent values.

```{r, fig.height=15, fig.width=10}
ggplot(candy) +
  aes(winpercent, rownames(candy)) + 
  geom_col()
```

> Q16. This is quite ugly, use the reorder() function to get the bars sorted by winpercent?

```{r, fig.height=15, fig.width=10}
ggplot(candy) +
  aes(winpercent, reorder(rownames(candy),winpercent)) + 
  geom_col()
```

Adding some color

```{r, fig.height=15, fig.width=10}
mycols <- rep("gray", nrow(candy))

## Chocolate candy
mycols[as.logical(candy$chocolate)] <- "brown"

## Fruity candy
mycols[as.logical(candy$fruity)] <- "pink"

## Bar candy
mycols[as.logical(candy$bar)] <- "cyan3"

ggplot(candy) + 
  aes(winpercent, reorder(rownames(candy),winpercent)) +
  geom_col(fill = mycols)
```

> Q17. What is the worst ranked chocolate candy?
> 
> > Sixlets

> Q18. What is the best ranked fruity candy?
> 
> > Starburst


## Taking a look at pricepercent

> Q19. Which candy type is the highest ranked in terms of winpercent for the least money - i.e. offers the most bang for your buck?

```{r}
ggplot(candy) +
  aes(winpercent, pricepercent, label=rownames(candy)) +
  geom_point(col=mycols) +
  geom_text()
```

We can fix the label text overplotting with an add-on package called **ggrepel** and it's `geom_text_repel()`

```{r}
library(ggrepel)

ggplot(candy) +
  aes(winpercent, pricepercent, label=rownames(candy)) +
  geom_point(col=mycols) +
  geom_text_repel(col=mycols, size=3, max.overlaps = 6)
```

> Q20. What are the top 5 most expensive candy types in the dataset and of these which is the least popular?
>
> > Nik L Nip

```{r}
candy |>
  arrange(desc(pricepercent)) |> 
  select(pricepercent, winpercent) |> 
  head(n=5)
```

## Exploring the correlation structure

We can calculate the pair-wise correlation of all our columns

```{r}
cij <- cor(candy)
cij
```

```{r}
library(corrplot)
corrplot(cij)
```

> Q22. Examining this plot what two variables are anti-correlated (i.e. have minus values)?
>
> > fruity and chocolate

> Q23. Use your corrplot result to make a prediction: which variables do you expect will have the largest contributions (i.e. loadings) to PC1 (i.e., drive the most separation between candies along the first principal component)?
>
> > fruity and chocolate

## Principal Component Analysis

```{r}
pca <- prcomp(candy, scale=TRUE)
summary(pca)
```

First major result figure is the "score plot" of PC1 vs PC2 - how the different candy are related to each other on our new PC axis:

```{r}
my_data <- cbind(candy, pca$x[,1:3])
p <- ggplot(my_data) +
  aes(PC1, PC2, size=winpercent/100, label=row.names(my_data)) +
  geom_point(col=mycols) +
  geom_text_repel(col=mycols, size=2)

p
```

```{r}
library(plotly)
#ggplotly(p)
```

The second major results figure from PCA

```{r}
ggplot(pca$rotation) +
  aes(PC1, reorder(row.names(pca$rotation), PC1)) +
  geom_col()
```

> Q24. Complete the code to generate the loadings plot above. What original variables are picked up strongly by PC1 in the positive direction? Do these make sense to you? Where did you see this relationship highlighted previously?
> 
> > fruity, pluribus, and hard are picked up strongly by PC1 in the positive direction. These make sense as this relationship was highlighted by corrplot previously.

> Q25. Based on your exploratory analysis, correlation findings, and PCA results, what combination of characteristics appears to make a “winning” candy? How do these different analyses (visualization, correlation, PCA) support or complement each other in reaching this conclusion?
> 
> > chocolate candy without fruity flavor (supported by PCA barplot and correlation plot)