---
title: "hw12"
author: "Julianne Lee (PID: A17979401)"
format: pdf
---

> Q13. Read this file into R and determine the sample size for each genotype and their
corresponding median expression levels for each of these genotypes.

```{r}
data <- read.table("https://bioboot.github.io/bggn213_W19/class-material/rs8067378_ENSG00000172057.6.txt", header = TRUE)

head(data)
```

```{r}
summary(data)
```

```{r}
table(data$geno)
tapply(data$exp, data$geno, median)
```

> Q14. Generate a boxplot with a box per genotype, what could you infer from the relative expression value between A/A and G/G displayed in this plot? Does the SNP effect the expression of ORMDL3?

```{r}
library(ggplot2)

ggplot(data, aes(x = geno, y = exp, fill = geno)) +
  geom_boxplot() +
  labs(
    title = "ORMDL3 Expression by rs8067378 Genotype",
    x = "Genotype",
    y = "ORMDL3 Expression"
  ) +
  theme_minimal()
```

> > ORMDL3 expression order is A/A, A/G, and G/G, from highest to lowest. This suggests that rs8067378 is associated with ORMDL3 expression, presence of G allele being correlated to lower expression levels. Therefore, the SNP affects the expression of ORMDL3.