---
title: "Class 13: Transcriptomics and RNA-seq Analysis"
author: "Julianne Lee (PID: A17979401)"
format: pdf
toc: true
---

## Bioconductor setup

```{r}
library(BiocManager)
library(DESeq2)
```

## Import countData and colData

```{r}
counts <- read.csv("airway_scaledcounts.csv", row.names = 1)
metadata <- read.csv("airway_metadata.csv")

head(counts)
head(metadata)
```

> Q1. How many genes are in this dataset?

```{r}
nrow(counts)
```

> Q2. How many ‘control’ cell lines do we have? 

```{r}
sum(metadata$dex == "control")
```

## Toy differential gene expression

```{r}
control <- metadata[metadata[,"dex"]=="control",]
control.counts <- counts[ ,control$id]
control.mean <- rowSums( control.counts )/4 
head(control.mean)
```

```{r}
library(dplyr)
control <- metadata %>% filter(dex=="control")
control.counts <- counts %>% select(control$id) 
control.mean <- rowSums(control.counts)/4
head(control.mean)
```

> Q3. How would you make the above code in either approach more robust? Is there a function that could help here?

```{r}
control <- metadata[metadata[,"dex"]=="control",]
control.counts <- counts[ ,control$id]
control.mean <- rowMeans(control.counts)
head(control.mean)
```

> Q4. Follow the same procedure for the treated samples (i.e. calculate the mean per gene across drug treated samples and assign to a labeled vector called treated.mean)

```{r}
treated <- metadata[metadata$dex == "treated", ]
treated.counts <- counts[, treated$id]
treated.mean <- rowMeans(treated.counts)

head(treated.mean)
```

```{r}
meancounts <- data.frame(control.mean, treated.mean)
```

> Q5 (a). Create a scatter plot showing the mean of the treated samples against the mean of the control samples. Your plot should look something like the following.

```{r}
meancounts <- data.frame(control.mean, treated.mean)
head(meancounts)
colSums(meancounts)
```

```{r}
plot(meancounts[,1], meancounts[,2],
     xlab = "Control",
     ylab = "Treated")                              
```

> Q5 (b).You could also use the ggplot2 package to make this figure producing the plot below. What geom_?() function would you use for this plot?

```{r}
library(ggplot2)

ggplot(meancounts, aes(x = control.mean, y = treated.mean)) +
  geom_point()
```

> Q6. Try plotting both axes on a log scale. What is the argument to plot() that allows you to do this?

```{r}
plot(meancounts$control.mean, meancounts$treated.mean,
     log = "xy",
     xlab = "Control",
     ylab = "Treated")
```

```{r}
meancounts$log2fc <- log2(meancounts[,"treated.mean"]/meancounts[,"control.mean"])
head(meancounts)
```

```{r}
zero.vals <- which(meancounts[,1:2]==0, arr.ind=TRUE)

to.rm <- unique(zero.vals[,1])
mycounts <- meancounts[-to.rm,]
head(mycounts)
```

> Q7. What is the purpose of the arr.ind argument in the which() function call above? Why would we then take the first column of the output and need to call the unique() function?

```{r}
zero.vals <- which(meancounts[, 1:2] == 0, arr.ind = TRUE)
#zero.vals
```

> > `arr.ind` makes `which()` to return the row and column positions of the values that match the condition. We take first column because the first column contains the gene into to be removed.

> Q8. Using the up.ind vector above can you determine how many up regulated genes we have at the greater than 2 fc level? 

```{r}
up.ind <- mycounts$log2fc > 2
sum(up.ind)
```

> Q9. Using the down.ind vector above can you determine how many down regulated genes we have at the greater than 2 fc level? 

```{r}
down.ind <- mycounts$log2fc < -2
sum(down.ind)
```

> Q10. Do you trust these results? Why or why not?
>
> > No, since this uses log2 fold change. The result can be unreliable if the read counts are low or highly variable. Also, this is not performing statistical analysis.

## Setting up for DESeq

```{r}
citation("DESeq2")
```

```{r}
dds <- DESeqDataSetFromMatrix(countData=counts, 
                              colData=metadata, 
                              design=~dex)
dds
```

## Principal Component Analysis (PCA)

```{r}
vsd <- vst(dds, blind = FALSE)
plotPCA(vsd, intgroup = c("dex"))
```

```{r}
pcaData <- plotPCA(vsd, intgroup=c("dex"), returnData=TRUE)
head(pcaData)
```

```{r}
percentVar <- round(100 * attr(pcaData, "percentVar"))
```

```{r}
ggplot(pcaData) +
  aes(x = PC1, y = PC2, color = dex) +
  geom_point(size =3) +
  xlab(paste0("PC1: ", percentVar[1], "% variance")) +
  ylab(paste0("PC2: ", percentVar[2], "% variance")) +
  coord_fixed() +
  theme_bw()
```

## DESeq analysis

```{r}
dds <- DESeq(dds)
res <- results(dds)
res
```

```{r}
summary(res)
```

```{r}
res05 <- results(dds, alpha=0.05)
summary(res05)
```

## Adding annotation data
```{r}
#BiocManager::install("AnnotationDbi")
#BiocManager::install("org.Hs.eg.db")
```

```{r}
library("AnnotationDbi")
library("org.Hs.eg.db")
```

```{r}
columns(org.Hs.eg.db)
```

```{r}
res$symbol <- mapIds(org.Hs.eg.db,
                     keys=row.names(res), 
                     keytype="ENSEMBL",  
                     column="SYMBOL", 
                     multiVals="first")
```

```{r}
head(res)
```

> Q11. Run the mapIds() function two more times to add the Entrez ID and UniProt accession and GENENAME as new columns called res$entrez, res$uniprot and res$genename.

```{r}
res$entrez <- mapIds(org.Hs.eg.db,
                     keys = row.names(res),
                     keytype = "ENSEMBL",
                     column = "ENTREZID",
                     multiVals = "first")

res$uniprot <- mapIds(org.Hs.eg.db,
                      keys = row.names(res),
                      keytype = "ENSEMBL",
                      column = "UNIPROT",
                      multiVals = "first")

res$genename <- mapIds(org.Hs.eg.db,
                       keys = row.names(res),
                       keytype = "ENSEMBL",
                       column = "GENENAME",
                       multiVals = "first")

head(res)
```

```{r}
ord <- order( res$padj )
head(res[ord,])
```

```{r}
write.csv(res[ord,], "deseq_results.csv")
```

## Data Visualization

```{r}
plot( res$log2FoldChange,  -log(res$padj), 
      xlab="Log2(FoldChange)",
      ylab="-Log(P-value)")
```

```{r}
plot( res$log2FoldChange,  -log(res$padj), 
 ylab="-Log(P-value)", xlab="Log2(FoldChange)")

abline(v=c(-2,2), col="darkgray", lty=2)
abline(h=-log(0.05), col="darkgray", lty=2)
```

```{r}
mycols <- rep("gray", nrow(res))
mycols[ abs(res$log2FoldChange) > 2 ]  <- "red"

inds <- (res$padj < 0.01) & (abs(res$log2FoldChange) > 2 )
mycols[ inds ] <- "blue"

plot( res$log2FoldChange,  -log(res$padj), 
 col=mycols, ylab="-Log(P-value)", xlab="Log2(FoldChange)" )

abline(v=c(-2,2), col="gray", lty=2)
abline(h=-log(0.1), col="gray", lty=2)
```

```{r}
#BiocManager::install("EnhancedVolcano")
```

```{r}
library(EnhancedVolcano)

x <- as.data.frame(res)

EnhancedVolcano(x,
    lab = x$symbol,
    x = 'log2FoldChange',
    y = 'pvalue')
```

