---
title: "hw_class06"
author: "Julianne Lee (A17979401)"
format: pdf
toc: TRUE
---

# Section 1

## A.

> Improve this regular R code by abstracting the main activities in your own new function.

```{r}
df <- data.frame(a=1:10, b=seq(200,400,length=10),c=11:20,d=NA)
df$a <- (df$a - min(df$a)) / (max(df$a) - min(df$a))
df$b <- (df$b - min(df$a)) / (max(df$b) - min(df$b))
df$c <- (df$c - min(df$c)) / (max(df$c) - min(df$c))
df$d <- (df$d - min(df$d)) / (max(df$a) - min(df$d)) 
```

Improvement

```{r}
dffunction <- function(x) {
  (x - min(x)) / (max(x) - min(x))
}
```

```{r}
df <- data.frame(a=1:10, b=seq(200,400,length=10),c=11:20,d=NA)
df$a <- dffunction(df$a)
df$b <- dffunction(df$b)
df$c <- dffunction(df$c)
df$d <- dffunction(df$d)

df
```

## B.

> Next improve the below example code for the analysis of protein drug interactions by abstracting the main activities in your own new function.

```{r}
library(bio3d)
s1 <- read.pdb("4AKE") # kinase with drug
s2 <- read.pdb("1AKE") # kinase no drug
s3 <- read.pdb("1E4Y") # kinase with drug
s1.chainA <- trim.pdb(s1, chain="A", elety="CA")
s2.chainA <- trim.pdb(s2, chain="A", elety="CA")
s3.chainA <- trim.pdb(s3, chain="A", elety="CA")
s1.b <- s1.chainA$atom$b
s2.b <- s2.chainA$atom$b
s3.b <- s3.chainA$atom$b
plotb3(s1.b, sse=s1.chainA, typ="l", ylab="Bfactor")
plotb3(s2.b, sse=s2.chainA, typ="l", ylab="Bfactor")
plotb3(s3.b, sse=s3.chainA, typ="l", ylab="Bfactor")
```

Improvement

```{r}
bio3dfunction <- function(id) {
  pdb <- read.pdb(id)
  pdb_chainA <- trim.pdb(pdb, chain="A", elety="CA")
  pdb_b <- pdb_chainA$atom$b
  plotb3(pdb_b, sse=pdb_chainA, typ="l", ylab="Bfactor")
}
```

```{r}
bio3dfunction("4AKE")
bio3dfunction("1AKE")
bio3dfunction("1E4Y")
```

## HW (Use code in B.)

> Q6. How would you generalize the original code above to work with any set of input protein structures?
>
> Write your own function starting from the code above that analyzes protein drug interactions by reading in any protein PDB data and outputs a plot for the specified protein.

```{r}
bio3dfunction <- function(id) { # INPUT: pdb id
  pdb <- read.pdb(id) # read pdb file of specified id
  pdb_chainA <- trim.pdb(pdb, chain="A", elety="CA") # stores ONLY chain A and CA atoms from pdb to 'pdb_chainA'
  pdb_b <- pdb_chainA$atom$b # stores numeric vector to 'pdb_b'
  plotb3(pdb_b, sse=pdb_chainA, typ="l", ylab="Bfactor") # plot B-factor using numeric vector 'pdb_b' with secondary structure elements into line plot, and add y-axis title as "Bfactor"
} # OUTPUT: final B-factor vs. Residue plot
```

```{r}
bio3dfunction("4AKE")
bio3dfunction("1AKE")
bio3dfunction("1E4Y")
```
