---
title: "Class 7: Machine Learning 1"
author: "Julianne Lee (PID: A17979401)"
format: pdf
toc: true
---

## Background

Today we will explore some core machine learning methods that are very populat in bioinformatics. These include **clustering** and **dimensionality reduction**.


## K-means clustering

The main function in "base" R for K-means clustering is called `kmeans()`

Before we go too deep let's make up some "simple" data that we can cluster and know if we are getting a good answer or not. To do this, we can use `rnomr()` function:

```{r}
hist(rnorm(10000, mean = 3))
```

```{r}
x <- c(rnorm(30, -3), rnorm(30, 3))
x
```

```{r}
z <- cbind(x, rev(x))
z
```

```{r}
plot(z)
```

```{r}
p <- 1:5

cbind(p, rev(p))
```

Now we can run `kmeans()` on this input `z` and see what the results look like. 

```{r}
kmeans(z, centers=2)
```

```{r}
km <- kmeans(z,centers=2)
km
```

```{r}
attributes(km)
```

> Q. How many points are in each cluster?

```{r}
km$size
```

> Q. What component of your result object details cluster assignment/memership?

```{r}
km$centers
```

> Q. Plot `z` colored by the kmeans cluster assignment and add cluster centers as blue points

```{r}
plot(z, col=km$cluster)+
  points(km$centers, col="blue", pch=16)
```

> Q. Run a K-means clustering and plot the results asking for 4 clusters (K=4)

```{r}
km2 <- kmeans(z, centers=4)
km2

plot(z,col=km2$cluster)+
  points(km2$centers, col="blue", pch=16)
```

> **N.B.** You need to tell K-means the number of clusters (i.e. set `centers=2`)!!

One approach is to try different values for `centers` and then pick the best...

```{r}
ans <- NULL
for (i in 1:10) {
  km <- kmeans(z, centers=i)
  ans <- c(ans,km$tot.withinss)
}

plot(ans,typ="o",
     xlab="Number of clusters",
     ylab="Total Sum of Squares Distance")
```

## Hierarchical Clustering

The main function in "base" R for Hierarchical Clustering is called `hclust()`

This function does not take your "raw" data for clustering. You must first build a "distance matrix" from your data and pass this as input to `hclust()`

```{r}
d <- dist(z)
hc <- hclust(d)
hc
```

There is a bespoke `plot()` method for `hclust()` result objects.

```{r}
plot(hc)+
abline(h=8, col="red")
```

Once we have our `hclust` object (our "tree" of "cluster dendrogram") we can *"cut"* the tree to reveal the clustering pattern.

```{r}
cutree(hc, k=4)
```

> Q. Make a plot of `z` with your hclust results (i.e. colored by cluster membership)

```{r}
grps <- cutree(hc, k=2)
plot(z, col=grps)
```

## Principal Component Analysis (PCA)

PCA is a dimensionality reduction method that is popular for revealing patterns in complex datasets.

### Analysis of UK food data

Let's look at some data on the eating habits of folks from the UK to see if there are patterns and trends that have some regions being distinct from others.

## Data import

The data is made available in CSV format so we can use the `read.csv()` function for import to R:

```{r}
url <- "https://tinyurl.com/UK-foods"
x <- read.csv(url)
x
```

## Tidy the data

Fix anything that went wrong with data import.

## Exporatory analysis

Make some plots to help make sense of obvious trends..

## PCA

> Q1. How many rows and columns are in your new data frame named x? What R functions could you use to answer this questions?

```{r}
dim(x)
head(x)
```

```{r}
rownames(x) <- x[,1]
x <- x[,-1]
head(x)
```

```{r}
dim(x)
```

```{r}
x <- read.csv(url, row.names=1)
head(x)
```

> Q2. Which approach to solving the ‘row-names problem’ mentioned above do you prefer and why? Is one approach more robust than another under certain circumstances?
>
> > I prefer the second since the first approach keeps deleting the first row name if we run the code again. 

```{r}
barplot(as.matrix(x), beside=T, col=rainbow(nrow(x)))
```

> Q3. Changing what optional argument in the above barplot() function results in the following plot?
>
>> beside=...

```{r}
barplot(as.matrix(x), beside=F, col=rainbow(nrow(x)))
```

```{r}
library(tidyr)

x_long <- x |> 
          tibble::rownames_to_column("Food") |> 
          pivot_longer(cols = -Food, 
                       names_to = "Country", 
                       values_to = "Consumption")

dim(x_long)
head(x_long)
```

```{r}
library(ggplot2)

ggplot(x_long) +
  aes(x = Country, y = Consumption, fill = Food) +
  geom_col(position = "dodge") +
  theme_minimal()
```

> Q4. Changing what optional argument in the above ggplot() code results in a stacked barplot figure?
>
>> geom_col(position=...)

```{r}
ggplot(x_long) +
  aes(x = Country, y = Consumption, fill = Food) +
  geom_col(position = "stack") +
  theme_minimal()
```

> Q5. We can use the pairs() function to generate all pairwise plots for our countries. Can you make sense of the following code and resulting figure? What does it mean if a given point lies on the diagonal for a given plot?

```{r}
pairs(x, col=rainbow(nrow(x)), pch=16)
```

```{r}
library(pheatmap)

pheatmap( as.matrix(x) )
```

> **Key-point:** Even relatively small datasets can prove challenging to interpret.

## PCA to the rescue

The main function in "base" R for PCA is called `prcomp()`. This function wants the "observations" to be rows and the "variables" to be columns. 

So here we need to take the transpose of our `x` input object.

```{r}
pca <- prcomp(t(x))
summary(pca)
```

The returned `pca` object has components that we can use to make our main result figures:

```{r}
attributes(pca)
```

The main result figure from this analysis is called a **"PC score plot"** (a.k.a. an "ordination plot" "PC plot" or "PC1 vs PC2 plot"). 

This plot shows how samples (in this case countries) relate to each other along our new PC axis.

```{r}
mycols <- c("orange", "red", "blue", "darkgreen")

ggplot(pca$x) +
  aes(x = PC1, y = PC2) +
  geom_point(col=mycols)


pca$x
```

```{r}
ggplot(pca$x) +
  aes(x = PC1, y = PC2, label=row.names(pca$x)) +
  geom_point(col=mycols) +
  geom_text(size=3, vjust=2, col=mycols)

pca$x
```

```{r}
ggplot(pca$rotation) +
  aes(PC1, reorder(row.names(pca$rotation), PC1))+
  geom_col()
```

> Q9. Generate a similar ‘loadings plot’ for PC2. What two food groups feature prominantely and what does PC2 maninly tell us about?
>
> > Soft drinks and alcoholic drinks are prominant. PC2 mainly tells about a beverage consumption among countries in how much they consume the soft drinks and alcoholic drinks relative to the other foods.

```{r}
ggplot(pca$rotation) +
  aes(PC2, reorder(row.names(pca$rotation), PC2))+
  geom_col()
```

